---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# defm: Fit and simulate discrete binary exponential family models

<!-- badges: start -->
<!-- badges: end -->

Discrete exponential family models (DEFM) have a long tradition with extensive
development rooted in exponential random graph models (ERGMs.) Applicable to 
any form o data that can be represented as binary arrays, DEFMs provide a way
to model jointly distributed binary variables.

This package, built on top of the C++ library [`barry`](https://github.com/USCbiostats/barry),
provides a computationally efficient implementation of this family of models.

## Installation

You can install the development version of defm like so:

``` r
# FILL THIS IN! HOW CAN PEOPLE INSTALL YOUR DEV PACKAGE?
```

# Examples

## Example 1: Four outcomes

In this example, we will simulate a dataset that contains 1,000 individuals with
four different outcomes. The outcomes, which are 0/1 vectors, will be modeled
as Markov processes of order one. Future states of this 0/1 vector
are a function of the previous point in time. The following lines
simulate the baseline data:

```{r data-simulation}
library(defm)

# Simulation parameters
set.seed(1231)
n   <- 5000L # Count of individuals
n_y <- 4L    # Number of dependent variables
n_x <- 2L    # Number of independent variables

# Simulating how many observations we will have per individuals
n_reps <- sample(3:10, n, replace = TRUE)

# Final number of rows in the data
n_t <- sum(n_reps)

# Simulating the data
Y <- matrix(as.integer(runif(n_y * n_t) < .1), ncol = n_y)
colnames(Y) <- paste0("y", 1:n_y - 1)
X <- matrix(rnorm(n_x * n_t), ncol = n_x)
colnames(X) <- paste0("x", 1:n_x - 1L)
id <- rep(1L:n, n_reps)
time <- unlist(sapply(n_reps, \(x) 1:x))
```

Here is a brief look at the data structure. Remember, we still have not actually
simulated data **WITH THE MODEL**.

```{r echo = FALSE}
knitr::kable(head(cbind(id, time, Y, X), n = 10), row.names = FALSE)
```


For this example, we will simulate a model with the following features:

- **Ones**: Baseline density (prevalence of ones).
- **Ones x Attr 2**: Same as before, but weighted by one of the covariates. (simil to fixed effect)
- **Transition** : And a transition structure, in particular `y0 -> (y0, y1)`, which will be represented
as a matrix in the form:

```
1 0 0 0
1 1 0 0
```

Here is the factory function:

```{r factory-function}
# Creating the model and adding a couple of terms
build_model <- function(id., Y., X., order. = 1, par. = par.) {
  
  d_model. <- new_defm(id., Y., X., order = order.)

  term_defm_ones(d_model.)
  term_defm_ones(d_model., 1)
  
  transition <- matrix(NA_integer_, nrow = order. + 1, ncol = ncol(Y.))
  transition[c(1,2,4)] <- 1
  
  term_defm_transition(d_model., transition)
  
  init_defm(d_model.)
  
  d_model.
  
}
```

With this factory function, we will use it to simulate some data with the same
dimensions of the original dataset. In this case, the parameters used for the simulation
will be:

- **Ones**: -2, i.e., low density,
- **Ones x Attr 2**: 2, yet correlated with covariate \# 2,
- **Transition** : 5, And a high chance of observing the transition `y0 -> (y0, y1)`

```{r simulation}
sim_par <- c(-2, 2, 5)
d_model <- build_model(id, Y, X, order = 1, par. = sim_par)
simulated_Y <- sim_defm(d_model, sim_par)
head(cbind(id, simulated_Y))
```


Now, let's see if we can recover the parameters using MLE:

```{r mle}
d_model_sim <- build_model(id, simulated_Y, X, order = 1, par. = sim_par)
ans <- defm_mle(d_model_sim)

summary(ans)
```

We can also see the counts

```{r example1-counts, echo=FALSE}
obs_stats <- get_stats(d_model_sim)
colnames(obs_stats) <- names(d_model_sim)
colnames(simulated_Y) <- paste0("y", 1:n_y - 1)
cbind(id, simulated_Y, X, obs_stats) |> 
  head(n = 10) |> knitr::kable(row.names = FALSE, digits = 2L)
```


## Example 2: A fun model

Let's try out making some patterns

```{r}
n   <- 20L
n_y <- 10L
id <- rep(1L, n)
Y <- matrix(0L, nrow = n, ncol = n_y)
Y[1] <- 1L
X <- cbind(1:n + 0.1)

d_model <- new_defm(id = id, Y = Y, X = X, order = 1)

for (i in (1:(n_y - 1) - 1)) {
  transition <- matrix(NA_integer_, nrow = 2, ncol = n_y)
  transition[c(1:4) + 2 * i] <- c(1,0,0,1)
  term_defm_transition(d_model, transition)
}

transition <- matrix(NA_integer_, nrow = 2, ncol = n_y)
transition[c(n_y * 2 - 1, n_y * 2, 1, 2)] <- c(1,0,0,1)
term_defm_transition(d_model, transition)

term_defm_ones(d_model)

init_defm(d_model)

Y_sim <-sim_defm(d_model, par = c(rep(100, n_y), -10))

stopifnot(all(Y_sim[1,] == Y[1,]))
```

Let's see how it looks like

```{r echo = FALSE}
library(Matrix)
Matrix::image(as(Y_sim, "dgCMatrix"))
```


# Acknowledgement

This work was supported by the Assistant Secretary of Defense for Health Affairs endorsed by the Department of Defense, through the Psychological Health/Traumatic Brain Injury Research Program Long-Term Impact of Military-Relevant Brain Injury Consortium (LIMBIC) Award/W81XWH-18-PH/TBIRP-LIMBIC under Award No. I01 RX003443. The U.S. Army Medical Research Acquisition Activity, 839 Chandler Street, Fort Detrick MD 21702-5014 is the awarding and administering acquisition office. Opinions, interpretations, conclusions and recommendations are those of the author and are not necessarily endorsed by the Department of Defense. Any opinions, findings, conclusions recommendations expressed in this publication are those of the author(s) and do not necessarily reflect the views of the U.S. Government, the U.S. Department of Veterans Affairs or the Department of Defense and no official endorsement should be inferred.

# Code of Conduct
  
  Please note that the defm project is released with a [Contributor Code of Conduct](https://contributor-covenant.org/version/2/0/CODE_OF_CONDUCT.html). By contributing to this project, you agree to abide by its terms.
